---
title: "Data loading"
output: html_document
date: "2025-10-31"
---

## TIRTLseq text file data structure

TIRTLseq data is kept in tab-separated text files (.tsv). The files may be 'gzipped' (.tsv.gz) to save space.<br>Output for each sample consists of 3 files:

- `<sample-name>_TIRTLoutput.tsv(.gz)`-- Paired TCRs -- a table of computationally paired alpha/beta TCRs
- `<sample-name>_pseudobulk_TRA.tsv(.gz)` -- Alpha chain "pseudo-bulk" data -- average read counts/frequencies of alpha-chains across all wells on a plate
- `<sample-name>_pseudobulk_TRB.tsv(.gz)` -- Beta chain "pseudo-bulk" data -- average read counts/frequencies of beta-chains across all wells on a plate

## Loading TIRTLseq Data in R

The `load_tirtlseq()` function loads paired TCR TIRTLseq data from all text files in a directory.

The function can also automatically assemble a metadata table if the files are named in a way that specifies the sample metadata.

Here, our files are named `(marker)_(timepoint)_(version)_ ... .tsv.gz`, e.g.
`cd4_tp1_v2_TIRTLoutput.tsv`.

```{r message=FALSE,warning=FALSE}
library(TIRTLtools)
library(dplyr)
folder = system.file("extdata/SJTRC_TIRTL_seq_longitudinal", package = "TIRTLtools")
dir(folder)
ts_data = load_tirtlseq(folder, meta_columns = c("marker", "timepoint", "version"), sep = "_")
## these files are named (marker)_(timepoint)_(version)_etc.tsv.gz
```

## The TIRTLseq data object

In R, TIRTLseq data is stored in a "list" which can contain any type of unstructured data.

The list contains two slots.

- `meta` - a data frame with sample metadata
- `data` - a list with data for each sample

```{r}
class(ts_data) ## list
names(ts_data) ## [1] "data" "meta"
```

### The metadata table

The `$meta` slot contains a data frame with columns for `sample_id` (determined from the filename) and
any columns you specified with the "meta_columns" argument. The `label` column combines all of the
metadata columns into one string.

If you did not specify "meta_columns" in the `load_tirtlseq()` call, then
the table will only contain the `sample_id` and `label` columns.

```{r}
ts_data$meta
```

### The data list

The `$data` slot is also a "list" that contains one object for each sample, named by its `sample_id`.

For each sample, `$data$[sample-name]` is a "list" of three data frames.

- `$data$[sample-name]$alpha` contains the pseudo-bulk data for the alpha chain
- `$data$[sample-name]$beta` contains the pseudo-bulk data for the beta chain
- `$data$[sample-name]$paired` contains alpha and beta chains that were paired computationally by our MAD-HYPE and T-SHELL algorithms.

```{r}
names(ts_data$data) ## [1] "cd4_tp1_v2" "cd4_tp2_v2" "cd4_tp3_v2" "cd8_tp1_v2" "cd8_tp2_v2" "cd8_tp3_v2"
class(ts_data$data$cd4_tp1_v2) ## list
names(ts_data$data$cd4_tp1_v2) ## [1] "alpha"  "beta"   "paired"
```

#### The pseudo-bulk data frames

In the TIRTLseq assay, one experiment is repeated hundreds of times in a multi-well plate.
Generally, an experiment is repeated in a half plate (192 wells) or a whole plate (384 wells).
Alpha and beta TCRs are sequenced simultaneously, but separately in each well.

The reads for each well are processed using [MiXCR](https://mixcr.com/), which calls V and J segments,
assembles the CDR3 nucleotide sequence, and generates its corresponding amino acid sequence.

For each clonotype defined by a unique CDR3 nucleotide sequence, the following summary metrics are reported:

- `readCount` - sum of read counts over all wells for the clonotype
- `readFraction` - the fraction of all reads attributed to the clonotype
- `sem` - standard error of the mean for the read fraction.
- `n_wells` - the number of wells in which a clonotype is observed
- `max_wells` - the number of wells used for the experiment (generally the same for all clonotypes)
- `readCount_max` - the maximum read count in any well for the clonotype
- `readCount_median` - the median read count for a clonotype

`$data$[sample-name]$alpha` and `$data$[sample-name]$beta` contain the pseudo-bulk data for alpha and beta chains, respectively.

```{r}
ts_data$data$cd4_tp1_v2$alpha

ts_data$data$cd4_tp1_v2$beta
```

#### The paired data frame

We use computational algorithms based on occurrence patterns ([MAD-HYPE algorithm](https://academic.oup.com/bioinformatics/article/35/8/1318/5095649)) or correlation of read fractions ([T-SHELL](https://www.biorxiv.org/content/10.1101/2024.09.16.613345v2)) of TCR chains across all wells to identify alpha and beta chains present in the same clone.

In general, the MAD-HYPE method works well for moderately frequent clones, but fails for the most frequent clones if they are found in all or almost all wells.
The T-SHELL algorithm works well for pairing these very frequent clones.

For each pair, we report:

##### V/J/CDR3 sequences
- `alpha_nuc_seq`, `alpha_nuc` - alpha CDR3 nucleotide sequence
- `beta_nuc_seq`, `beta_nuc` - beta CDR3 nucleotide sequence
- `alpha_beta` - alpha-beta pair concatenated nucleotide sequence
- `cdr3a` - alpha CDR3aa sequence
- `va` - alpha V gene
- `ja` - alpha J gene
- `cdr3b` - beta CDR3aa sequence
- `vb` - beta V gene
- `jb` - beta J gene

##### Alpha/beta chain occurence and overlap
- `wa` - number of wells where alpha is found
- `wb` -number of wells where beta is found
- `wi` - number of wells where alpha is found, but not beta
- `wj` - number of wells where beta is found, but not alpha
- `wij` - number of wells where both alpha and beta are found

##### Pairing method scores and metrics
- `method` - method used for pairing: T-SHELL or MAD-HYPE
- `r` - for T-SHELL, the Pearson correlation coefficient between alpha and beta frequencies
- `ts` - for T-SHELL, the t-statistic of the correlation
- `pval` - for T-SHELL, the p-value of the correlation
- `pval_adj` - for T-SHELL, the adjusted p-value of the correlation
- `score` - for MAD-HYPE, the score of the pairing (higher is better)
- `loss_a_frac` - the fraction of wells with lost alpha chain
- `loss_b_frac` - the fraction of wells with lost beta chain


```{r}
ts_data$data$cd4_tp1_v2$paired
```

Note: Some TCR pairs will be listed twice in the `$paired` data frame, once for the "MAD-HYPE" method and once for the "T-SHELL" method.

```{r}
ts_data$data$cd4_tp1_v2$paired %>% 
  filter(alpha_beta == "TGTGCCGTGAACGGAGGTAACGACTACAAGCTCAGCTTT_TGTGCCAGCTCACCAGAGCAGGGTCTGGCCCAGCATTTT") %>%
  select(alpha_nuc, beta_nuc, method, everything())
```

To see only unique TCR pairs, you may do the following:

```{r}
pairs = ts_data$data$cd4_tp1_v2$paired
pairs_unique = pairs[!duplicated(pairs$alpha_beta),]
pairs_unique
```



