---
title: "Manual T-SHELL for individual chains"
output:
  html_document:
    df_print: paged
---

The following functions are meant to be used to investigate the pairing of 
individual chains of interest.

This could be useful for multi-paired chains,
abundant chains that have unexpectedly failed to pair, or for chains that 
expanded between timepoints or are of interest for other experimental reasons.

It could also be useful for exploring possible alterations to the pairing algorithms
or for generating figures for sanity checks and to show confident pairing for
individual clones.

```{r}
pkgload::load_all("~/git/TIRTLtools")
```

First, we load already-paired data. If we look at the alpha pseudobulk, we can see here that there are many very abundant clones that were not paired.
The most abundant un-paired alpha clone is 15.9% of the repertoire.

```{r}
## Bleakly miH expanded sample -- had issues pairing most abundant clones
data = load_tirtlseq("~/git/Bleakley_miH/data/Bleakley_miH_project", 
                     samples = c("Bleakley_miH_4_1_D28_HA1_exp"), verbose = F)
data = data %>% identify_paired(verbose = F)

un_paired = data$data$Bleakley_miH_4_1_D28_HA1_exp$alpha %>% 
  filter(!is_paired) %>% 
  select(targetSequences, readCount, readFraction, everything())
un_paired
```

## Reading well data and writing to a convenient, fast-loading format

Well-level data comes in the form of a directory of text files (.tsv) with
read counts for TCRalpha and TCRbeta chains. Each text file has the chain and the
well in the file name. Reading up to 384 text files is slow and can take multiple
minutes, even when optimized, so after loading the first time, we write them
to a more convenient, fast-loading format.

This format consists of sparse matrices of read counts (wells x clones) for TCRalpha
and TCRbeta saved as .rds, an R-specific binary format. The matrices are "dgCMatrix",
a column-based format that allows for extremely fast extraction of any column
(read counts for one clone across all wells).

This only needs to be run once.
```{r}
# Note: "Bleakley_miH_4_1_D28_HA1_exp" is MVP175, UD-21, A1 to H24
dir_in1 = "/Volumes/thomagrp/home/common/TIRTL/tirtlSeq_data/MVP175_Bleakley_miH/clones_MVP175_Bleakley_miH/TCR_clones_UD21/"
dir_out1 = "/Volumes/thomagrp/home/common/nclark52/temp/MVP175_Bleakley_miH/Bleakley_miH_4_1_D28_HA1_exp/"
prefix1 = "Bleakley_miH_4_1_D28_HA1_exp"
wells1 = get_well_subset(row_range = 1:8, col_range = 1:24)

### run once
# write_well_data_to_binary(
#   folder_in = dir_in1, folder_out = dir_out1, prefix = prefix1, wells = wells1, 
#   in_file_type = ".tsv", out_file_type = ".rds", 
#   well_pos = 3, to_sparse_matrix = TRUE
#   )
```

Next, we load the well-level data and run `plot_tshell()` for the most abundant un-paired clone:

```{r}
well_data = load_well_counts_binary(dir_out1, prefix1, lazy = FALSE)
nuc_seq = un_paired$targetSequences[1] ## most abundant un-paired clone (15% of repertoire)
res = plot_tshell(well_data, nuc_seq, chain="alpha", interactive = F, 
                  plot_log_scale = F, n_plot = 9, loss_frac_cutoff = 0.5)
res$plots
```

```{r}
res$data$df_top_n
```

